VariantSpark: Population Scale Clustering of Genotype Information

### 2015-09-11 doc/publications/journal2014/paper.tex
comments

We hence developed a more focused purpose-built application in {\sc Spark} to perform genomic clustering of individuals. 
We utilise {\sc Spark}'s machine learning library, \mbox{MLlib}, and provide an interface from the standard variant data format, Variant Call Format (VCF)~\cite{1KG2012}, which opens up the application of MLlib's different machine learning algorithms a wide range of genotype-based analysis tasks. 
To demonstrate its capability, we cluster variant datasets from the 1000 Genomes Project~\cite{1KG2012} to determine population structure using the \kMeans{} clustering algorithm available in MLlib's. 
In the first section we benchmark the performance and accuracy of MLlib's implementations against Hadoop's Mahout as well as more traditional methods (R, Python) limiting the used data to chromosome 22.
In section two we demonstrate {\sc Spark}'s full capacity by seamlessly scaling from 1\% to 100\% of the human genome.
In section three we replicate the analysis by using XX genomes from the Personal Genome Project~\cite{Lunshof2013}.
In the last section we discuss the pipeline for visualising the resulting cluster.

@@

Clustering the samples is also fastest in \variantSpark{} (1min and 20secs), despite \variantSpark{} and ADAM both using {\sc Spark}'s MLlib \kMeans{} implementation, see Table~\ref{fivewaycomparison}.
% DENIS: how sure are we that this is not just a fluke, e.g. I typically repeat this 5 times and report the average and std/ste ...
% AIDAN: This result was consistent. But it was on the Azure cluster so I can't get std/ste..
We attribute the 30\% speedup over ADAM (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient.
Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise 8 cores. 
Despite this, both were an order of magnitude slower than \variantSpark{} and ADAM, with python requiring 11mins and 29secs and R requiring 7mins and 25secs.
Our Hadoop implementation performed the worst with 14mins and 23secs.
The substantially increased runtime is due the dataset (chromosome 22) being small enough to fit into memory. This enables Python and R to store the output from each \kMeans{} iteration in memory. Hadoop, however, writes the entire output of each iteration to disk.
Although this feature allows for greater scalability, as memory availability is less of an issue with huge datasets, the bottleneck introduced from disk IO increases processing time notably.
{\sc Spark} overcomes this limitation by performing in-memory caching of datasets. Caching virtually eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where at least a portion of the data can still be cached in memory.

@@

We run this comparison on our in-house Hadoop cluster using 64 cores, although this was limited to 40 cores with 100\% of the genome due to memory availability.
Pre-processing the VCF files takes approximately 12 minutes, 19 minutes, 27 minutes and 40 minutes for 20\%, 40\%, 60\% and 100\% of the genome, respectively.
In each case, the memory required does not exceed a modest 2GB per executor, even for 100\% of the genome. 
Clustering the data takes 70 minutes, 139 minutes, 213 minutes and 884 mins, respectively.
For clustering, the memory requirements increase approximately linearly with 6GB, 12GB, 17GB and 24GB, respectively, per executor. 
This substantial increase in memory required is a result of the increased dimensionality of the data causing the \kMeans{} centroids, which MLlib stores as dense vectors, requiring excessive memory.  
% DENIS: This conflicts with the 30 seconds gain over ADAM which you attribute to our clustering using sparse vectors,
% AIDAN: Only centroids are dense. Reworded to explain this
% TODO: is it farther ?
% AIDAN: I had to check, but I believe it is further? (http://www.quickanddirtytips.com/education/grammar/further-versus-farther)
While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.

%DENIS: do you see a change in ARI from chr22 to 20%, 40% ?
%AIDAN: not really..

@@

%AIDAN: Changed this section..
Again we cluster the individuals allowing 5 clusters and compare the assigned labels to the same number of self-reported `Race/ethnicity' labels (`White', `Hispanic or Latino', `Asian', `Black or African American' and `American Indian').
The observed ARI is negative, indicating the clustering is completely random. 
This poor result may be due to the labels being too broad and being a blanket label for a diverse range of individuals. For example, the `White' label includes individuals with grandparents from United States to Syrian Arab Republic to Bulgaria.
The dataset is also very one-sided with 433 out of the 478 (over 90\%) of the individuals being classified as white, even though geographical location of their grandparents are very diverse.
%This poor result may be due to the population annotation being too detailed and diverse, especially so for self-reported, and hence potential uncertain ancestry.
% TODO provide mapping
We therefore only include individuals where all grandparents were reported to be from the same country and group them into their approximate super-populations (see Supplemental Table X). 
This new dataset comprises 36 individuals.
Clustering these individuals results in an ARI of 0.29.

%DENIS: If you see an increase in accuracy above we can hypothesise here that 985,790 variants are not sufficient... otherwise a sentence around the population mix in potential real world clinical cohorts.... 
Although we see an increase, the relatively small ARI is likely due to the diversity in this real world clinical cohort across the small sample size. Although our filter removes individuals with a mixed background, we are left with only 36 individuals from 23 different countries.
Applying a super-population mask, based on geographical distance, to these individuals reduces the number of groups from 23 to 5, however this grouping doesn't take migrations that occurred prior to their grandparents into account.

@@

\section*{Conclusions}
%DENIS: mention genome wide sampling ability

\variantSpark{} excels at pre-processing VCF files into row vectors, being over an order of magnitude faster than ADAM. Memory consumption is also kept to a minimum. Where traditional approaches such as Python and R would fail when expanding to the whole genome,
\variantSpark{} can pre-process VCF files with over 1000 individuals in under half an hour using minimal memory.

Clustering these individuals is also much faster with \variantSpark{} than Python and R, although only marginally faster than ADAM.

These benefits of speed, resource consumption and scalability demonstrate \variantSpark{} to be a useful tool in applying machine learning tools to genomic data. Although we only run \kMeans on the data, MLlib provides many other machine learning algorithms, such as Random Forests.
Therefore it may be useful to investigate the novelty of applying these machine learning approaches to big genomic data.

>>>>

We hence developed a more focused purpose-built application in {\sc Spark} to perform genomic clustering of individuals. 
We utilise {\sc Spark}'s machine learning library, \mbox{MLlib}, and provide an interface from the standard variant data format, Variant Call Format (VCF)~\cite{1KG2012}, which opens up the application of MLlib's different machine learning algorithms a wide range of genotype-based analysis tasks. 
To demonstrate its capability, we cluster variant datasets from the 1000 Genomes Project~\cite{1KG2012} to determine population structure using the \kMeans{} clustering algorithm available in MLlib's. 
In the first section we benchmark the performance and accuracy of MLlib's implementations against Hadoop's Mahout as well as more traditional methods (R, Python) limiting the used data to chromosome 22.
% TODO fill in
In section two we demonstrate {\sc Spark}'s full capacity by seamlessly scaling from 1\% to 100\% of the human genome.
In section three we replicate the analysis by using XX genomes from the Personal Genome Project~\cite{Lunshof2013}.
In the last section we discuss the pipeline for visualising the resulting cluster.

@@

Clustering the samples is also fastest in \variantSpark{} (1min and 20secs), despite \variantSpark{} and ADAM both using {\sc Spark}'s MLlib \kMeans{} implementation, see Table~\ref{fivewaycomparison}. % 80
% DENIS: how sure are we that this is not just a fluke, e.g. I typically repeat this 5 times and report the average and std/ste ...
% AIDAN: This result was consistent. But it was on the Azure cluster so I can't get std/ste..
% DENIS: OK but I thought all of them were run on the same machine and you are still generating results for python ... in any case can you write something like (repeated X times)
We attribute the 30\% speedup over ADAM (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient. %
Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise 8 cores. 
Despite this, both were an order of magnitude slower than \variantSpark{} and ADAM, with python requiring 11mins and 29secs and R requiring 7mins and 25secs.
Our Hadoop implementation performed the worst with 14mins and 23secs. % 863
The substantially increased runtime is due the dataset (chromosome 22) being small enough to fit into memory. This enables Python and R to store the output from each \kMeans{} iteration in memory. Hadoop, however, writes the entire output of each iteration to disk.
Although this feature allows for greater scalability, as memory availability is less of an issue with huge datasets, the bottleneck introduced from disk IO increases processing time notably.
{\sc Spark} overcomes this limitation by performing in-memory caching of datasets. Caching virtually eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where at least a portion of the data can still be cached in memory.

@@

We run this comparison on our in-house Hadoop cluster using 40 cores.
%We run this comparison on our in-house Hadoop cluster using 64 cores, although this was limited to 40 cores with 100\% of the genome due to memory availability.
Pre-processing the VCF files takes approximately 12 minutes, 19 minutes, 27 minutes and 40 minutes for 20\%, 40\%, 60\% and 100\% of the genome, respectively.
In each case, the memory required does not exceed a modest 2GB per executor, even for 100\% of the genome. 
Clustering the data takes 70 minutes, 139 minutes, 213 minutes and 884 mins, respectively.
For clustering, the memory requirements increase approximately linearly with 6GB, 12GB, 17GB and 24GB, respectively, per executor. 
This substantial increase in memory is a result of more \kMeans{} centroids being created, which MLlib stores as dense vectors. 
%DENIS: do you see a change in ARI from chr22 to 20%, 40% ?
%AIDAN: not really.. 
%DENIS: so it is still exactly 0.84? Can you provide the number of Variants used in the different sets (e.g. is 20\% less than chrom 22)
%TODO A citation wold be good here
We do not observe an increase in cluster accuracy when providing variants from other chromosomes indicating that current practice of clustering based on a subset of the genome is sufficient to define the population boundaries~\cite{Hunter2014}.

%This substantial increase in memory required is a result of the increased dimensionality of the data causing the \kMeans{} centroids, which MLlib stores as dense vectors, requiring excessive memory.  
% DENIS: This conflicts with the 30 seconds gain over ADAM which you attribute to our clustering using sparse vectors,
% AIDAN: Only centroids are dense. Reworded to explain this
% DENIS: That means you allow more clusters ?
While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.

@@

Again we cluster the individuals allowing 5 clusters and compare the assigned labels to the same number of self-reported `Race/ethnicity' labels (`White', `Hispanic or Latino', `Asian', `Black or African American' and `American Indian').
The observed ARI is negative, indicating the clustering is completely random. 
This poor result may be due to the labels being too broad and not accurately reflecting genetic diversity. For example, the `White' label includes individuals with grandparents from United States, Syria, Arab Republic, or Bulgaria, amongst others. 
%This poor result may be due to the labels being too broad and being a blanket label for a diverse range of individuals. For example, the `White' label includes individuals with grandparents from United States to Syrian Arab Republic to Bulgaria.
The dataset is also very one-sided with 433 out of the 478 (over 90\%) of the individuals being classified as white, even though geographical location of their grandparents are very diverse.
% TODO provide mapping
We therefore only include individuals where all grandparents were reported to be from the same country leaving 36 individuals from 23 different countries, which we group into their approximate super-populations (see Supplemental Table X). 
Clustering these individuals results in an ARI of 0.29.

Although we see an increase in cluster-accuracy when removing individuals with a mixed background and operating at super-population level, the clustering is not as precise as on the carefully curated and characterised individuals from the 1000 Genomes Project.
%TODO add numbers
This is especially the case since XX\% are 'White' Americans, which formed the group most difficult to cluster in the 1000 Genomes Project data.

This highlights the issue for clinical application where ancestries influences treatment (e.g. HLA allele genotyping from SNP information~\cite{Zheng2014}) and accurate population association may not be known for patients with diverse migrational background.
Hence fast genome-wide applicable approaches may help in providing the necessary depth to provide insight into genetic diversity.

%Although we see an increase, the relatively small ARI is likely due to the diversity in this real world clinical cohort across the small sample size. Although our filter removes individuals with a mixed background, we are left with only 36 individuals from 23 different countries.
%Applying a super-population mask, based on geographical distance, to these individuals reduces the number of groups from 23 to 5, however this grouping doesn't take migrations that occurred prior to their grandparents into account.

@@

\section*{Conclusions}

\variantSpark{} performs clustering on VCF files with over 1000 individuals and 40 Million variants in under 15 hours using minimal memory (24GB). 
\variantSpark{} supports random genome-wide sampling of variants allowing faster clustering for well characterised cohorts where 20\% of the genome is sufficient. 
On the benchmarking dataset, it outperforms ADAM by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
\variantSpark{} utilizes {\sc Spark}, which allows for in-memory cashing and hence performs 85\% faster than Hadoop (29 min), 94\% faster than R (75 min), and 95\% faster than Python (103 min).

%\variantSpark{} excels at pre-processing VCF files into row vectors, being over an order of magnitude faster than ADAM. Memory consumption is also kept to a minimum. Where traditional approaches such as Python and R would fail when expanding to the whole genome,
%\variantSpark{} can pre-process VCF files with over 1000 individuals in under half an hour using minimal memory.

%Clustering these individuals is also much faster with \variantSpark{} than Python and R, although only marginally faster than ADAM.

%These benefits of speed, resource consumption and scalability demonstrate \variantSpark{} to be a useful tool in applying machine learning tools to genomic data. Although we only run \kMeans on the data, MLlib provides many other machine learning algorithms, such as Random Forests.
%Therefore it may be useful to investigate the novelty of applying these machine learning approaches to big genomic data.
